---
title: 'Topics in Labor & Demo: Midterm Problem Set'
author: "Gabriel Tabak, Guilherme Cabrine & Gustavo Henrique"
date: '`r Sys.Date()`'
output: pdf_document
encoding: UTF-8
---


```{r, echo=FALSE}
## Loading/Installing needed packages
pacman::p_load(tidyverse,
               estimatr,
               stats,
               stargazer,
               haven,
               here,
               patchwork,
               readxl,
               xtable,
               kableExtra)
```

# Question 1

```{r, echo=FALSE}
## Importing databases
# Importing SDEM final dataset
sdem = read_csv(here("Data/Final", "sdem.csv"))

# Cleaning memory
gc()

# Price index Data
mexcpi = read_excel(here("Data/Raw", "MEXCPIALLAINMEI.xlsx"),
                     sheet = "Annual") %>% 
  mutate(year = as.numeric(format(observation_date, "%Y"))) %>% 
  rename(cpi = MEXCPIALLAINMEI) %>% 
  select(-1)
```

### Item a
```{r, echo=FALSE}
# Get some household variables 
sdem_agg = sdem %>% 
  group_by(cd_a, ent, con, v_sel, n_hog, h_mud, year) %>%
  summarise(age_hhead = eda[which(par_c == 101)],
            age_hspouse = first(eda[par_c %in% c(201,202)]),
            chld_less16 = as.integer(sum(par_c %in% c(301,302,303) &
                                           eda < 16) > 0))

# Join with the data
sdem = sdem %>% 
  left_join(sdem_agg)

# Removing unneeded df
rm(sdem_agg)

## Employed 
# hrsocu > 0 and non empty, also, age_hhead > 13 and age_hhead <= 98
sdem = sdem %>%
  mutate(Employed = case_when(
    (hrsocup > 0) & (age_hhead <= 98) & (age_hhead > 13) ~ 1,
    T ~ 0
      ),
    # Get the conditional hours
    HourC = case_when(Employed == 1 ~ hrsocup, 
                    T ~ NA
      )
    )

# Deflate the salary
sdem = sdem %>% 
  left_join(mexcpi, by = c("year"))

sdem = sdem %>%
  mutate(IngOcupYear = 12*ingocup*(100/cpi),
         Wage = 52*ing_x_hrs*hrsocup*(100/cpi))

df1 = sdem %>% 
  filter(sex == 2, 
         eda %in% 25:65) %>% 
  group_by(year) %>%
  summarise(EmplR = mean(Employed),
            YWage = mean(Wage),
            Hrs = mean(hrsocup),
            HrsC = mean(HourC, na.rm = T)) %>% 
  mutate(year = as.factor(year))

g1 = ggplot(df1, aes(year, EmplR)) + geom_point() + 
  geom_line(aes(group = 1)) + theme_bw() +
  ylab("Employment")

g2 = ggplot(df1, aes(year, YWage)) + geom_point() + 
  geom_line(aes(group = 1)) +theme_bw() +
  ylab("Mean yearly Wage")

g3 = ggplot(df1, aes(year, Hrs)) + geom_point() + 
  geom_line(aes(group = 1)) + theme_bw() +
  ylab("Hours Occupied by week Unconditional")

g4 = ggplot(df1, aes(year, HrsC)) + geom_point() + 
  geom_line(aes(group = 1)) + theme_bw() +
  ylab("Hours Occupied by week Conditional")

# Combining graphics with dashboard identification
combined_plot = (g1 | g2) / (g3 | g4) +
  plot_annotation(tag_levels = 'A')

# Saving the combined graph
ggsave(here("Out", "plot1.png"), combined_plot, width = 12, height = 12, dpi=600)

rm(g1, g2, g3, g4, df1, combined_plot)

gc()
```

### Item b
```{r, echo=FALSE}
# DATA PREPARATION SECTION
# Filter the main dataset (sdem) to include only women aged 25-65 and create child indicator
df = sdem %>%
  filter(sex == 2, eda %in% 25:65) %>%  # Filter for females (sex == 2) aged 25-65
  mutate(hj_peq = case_when(
    # Create indicator for women with young children: 1 if has children AND has child under 16 AND is in partnership
    n_hij != 0 & chld_less16 != 0 & (par_c %in% c(101,201)) ~ 1,
    T ~ 0  # Otherwise assign 0 (no young children or not in partnership)
  ))

# MAIN TABLE GENERATION FUNCTION
# Function to create summary statistics tables by marital status
Tables = function(df1, status){
  # Filter dataset by specific marital status (5=married, 1=cohabiting, 6=single)
  df1 = df1 %>% 
    filter(e_con == status)
  
  # Initialize empty 5x8 matrix to store results (5 age groups, 8 columns for stats)
  final = matrix(0, 5, 8)
  final = data.frame(final)
  
  # ROW 1: ALL WOMEN (regardless of age/children status)
  # Columns 1&5: Overall employment rate and weekly earnings for all women
  final[1,c(1,5)] = c(mean(df1$Employed), mean(df1$Wage/52))
  
  # Columns 2&6: Employment rate and weekly earnings for women with less than high school education
  final[1,c(2,6)] = df1 %>% 
    filter(cs_p13_1 < 4) %>%  # Education level less than 4 (less than HS)
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  # Columns 3&7: Employment rate and weekly earnings for women with high school education
  final[1,c(3,7)] = df1 %>% 
    filter(cs_p13_1 == 4) %>%  # Education level equals 4 (high school)
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  # Columns 4&8: Employment rate and weekly earnings for women with college or above
  final[1,c(4,8)] = df1 %>%
    filter(cs_p13_1 %in% c(7,8,9)) %>%  # Education levels 7,8,9 (college or above)
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  # ROW 2: WOMEN UNDER 35 WITH YOUNG CHILDREN
  # Overall stats for women under 35 with young children
  final[2,c(1,5)] = df1 %>% 
    filter(hj_peq == 1, eda < 35) %>%  # Has young child indicator = 1 AND age < 35
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  # Same group broken down by education levels (less than HS)
  final[2,c(2,6)] = df1 %>% 
    filter(hj_peq == 1, eda < 35, cs_p13_1 < 4) %>% 
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  # Same group broken down by education levels (high school)
  final[2,c(3,7)] = df1 %>% 
    filter(hj_peq == 1, eda < 35, cs_p13_1 == 4) %>% 
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  # Same group broken down by education levels (college+)
  final[2,c(4,8)] = df1 %>% 
    filter(hj_peq == 1, eda < 35,cs_p13_1 %in% c(7,8,9)) %>%
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  # ROW 3: WOMEN UNDER 35 WITH NO SMALL CHILDREN
  # Overall stats for women under 35 without young children
  final[3,c(1,5)] = df1 %>% 
    filter(hj_peq == 0, eda < 35) %>%  # No young child indicator = 0 AND age < 35
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  # Same group broken down by education levels (less than HS, HS, college+)
  final[3,c(2,6)] = df1 %>%
    filter(hj_peq == 0, eda < 35,cs_p13_1 < 4) %>%
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  final[3,c(3,7)] = df1 %>%
    filter(hj_peq == 0, eda < 35,cs_p13_1 == 4) %>%
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  final[3,c(4,8)] = df1 %>% 
    filter(hj_peq == 0, eda < 35,cs_p13_1 %in% c(7,8,9)) %>%
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  # ROW 4: WOMEN AGED 35-54
  # Overall stats for middle-aged women (35-54)
  final[4,c(1,5)] = df1 %>% 
    filter(eda >= 35, eda<=54) %>%  # Age between 35 and 54
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  # Same group broken down by education levels
  final[4,c(2,6)] = df1 %>%
    filter(eda >= 35, eda<=54,cs_p13_1 < 4) %>% 
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  final[4,c(3,7)] = df1 %>%
    filter(eda >= 35, eda<=54,cs_p13_1 == 4) %>%
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  final[4,c(4,8)] = df1 %>%
    filter(eda >= 35, eda<=54,cs_p13_1 %in% c(7,8,9)) %>%
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  # ROW 5: WOMEN AGED 55 AND OLDER
  # Overall stats for older women (55+)
  final[5,c(1,5)] = df1 %>% 
    filter(eda >= 55) %>%  # Age 55 and above
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  # Same group broken down by education levels
  final[5,c(2,6)] = df1 %>% 
    filter(eda >= 55,cs_p13_1 < 4) %>% 
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  final[5,c(3,7)] = df1 %>% 
    filter(eda >= 55,cs_p13_1 == 4) %>%
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  final[5,c(4,8)] = df1 %>% 
    filter(eda >= 55,cs_p13_1 %in% c(7,8,9)) %>%
    summarise(mean(Employed), mean(Wage/52)) %>%
    as.numeric()
  
  # Assign descriptive row names for the 5 age/family status groups
  rownames(final) = c("All", "Less than 35 with Child Younger than 16",
                      "Less than 35 with No Small Children", "35-54",
                      "55 and Older")
  
  # Assign column names: first 4 columns for labor force participation, last 4 for earnings
  colnames(final) = c("All", "Less than HS", "High School", "College or Above",
                      "All", "Less than HS", "High School", "College or Above")  
  return(final)
}

# LATEX TABLE GENERATION FUNCTION
# Function to convert data tables into professionally formatted LaTeX tables
create_final_table = function(df, status, status_name, output_file = NULL) {
  # Generate the data table using the Tables function
  table_data = Tables(df, status)
  
  # Create a clean label name for LaTeX referencing (lowercase, hyphens instead of spaces)
  label_name = tolower(gsub(" ", "-", status_name))
  
  # Build LaTeX table code following professional academic formatting standards
  latex_code = c(
    "\\begin{table}[htbp]",  # Table environment with positioning options
    "    \\centering",        # Center the table
    paste("    \\caption{", status_name, "Aged 25-65}"),  # Table caption
    paste("    \\label{tab:", label_name, "}", sep = ""),  # Table label for cross-referencing
    "    \\resizebox{\\textwidth}{!}{%",  # Resize table to fit page width
    "    \\begin{tabular}{lcccccccc}",    # Table structure: 1 left-aligned, 8 center-aligned columns
    "    \\toprule",  # Top horizontal line (booktabs package)
    # Create grouped column headers spanning multiple columns
    "     & \\multicolumn{4}{c}{\\textbf{Labor Force Participation}} & \\multicolumn{4}{c}{\\textbf{Earnings}} \\\\",
    "    \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}",  # Partial horizontal lines under group headers
    # Individual column headers
    "     & All & Less than HS & High School & College+ & All & Less than HS & High School & College+ \\\\",
    "    \\midrule"  # Middle horizontal line separating headers from data
  )
  
  # Define row labels matching the data structure
  row_names = c("All", 
                "Less than 35 with Child Younger than 16",
                "Less than 35 with No Small Children", 
                "35-54",
                "55 and Older")
  
  # Add data rows to LaTeX code
  for(i in 1:nrow(table_data)) {
    # Format numbers to 3 decimal places for consistent presentation
    formatted_numbers = sprintf("%.3f", table_data[i,])
    # Construct each data row with proper LaTeX formatting
    data_row = paste("    ", row_names[i], " & ", 
                     paste(formatted_numbers, collapse = " & "),
                     " \\\\", sep = "")
    latex_code = c(latex_code, data_row)
  }
  
  # Close the LaTeX table structure
  latex_code = c(latex_code,
                 "    \\bottomrule",      # Bottom horizontal line
                 "    \\end{tabular}%",  # End tabular environment
                 "    }",                # Close resizebox
                 "    \\end{table}")     # End table environment
  
  # Output handling: save to file if specified, always print to console
  if(!is.null(output_file)) {
    writeLines(latex_code, output_file)  # Write LaTeX code to .tex file
  }
  
  cat(latex_code, sep = "\n")  # Display LaTeX code in console
  return(latex_code)           # Return code for further processing if needed
}

# BATCH TABLE GENERATION FUNCTION
# Utility function to generate all three required tables at once
generate_all_tables = function(df) {
  # Define the three marital status categories as specified in the problem set
  status_list = list(
    list(code = 5, name = "Married Women"),      # Marital status code 5
    list(code = 1, name = "Cohabiting Women"),   # Marital status code 1
    list(code = 6, name = "Single Women")        # Marital status code 6
  )
  
  # Loop through each marital status and generate corresponding table
  for(status_info in status_list) {
    # Print progress indicators for user feedback
    cat("\n", rep("=", 60), "\n")
    cat("GENERATING TABLE FOR:", status_info$name, "\n")
    cat(rep("=", 60), "\n")
    
    # Create standardized filename based on group name
    filename = paste0(tolower(gsub(" ", "_", status_info$name)), "_table.tex")
    # Generate the table
    create_final_table(df, status_info$code, status_info$name, filename)
    
    cat("\nTable saved to:", filename, "\n")
  }
}

# EXECUTION SECTION
# Generate all tables using the batch function
generate_all_tables(df)

# Alternative: Generate tables individually with specific file paths (using 'here' package for path management)
create_final_table(df, 5, "Married Women", here("Out", "married_women_final.tex"))
create_final_table(df, 1, "Cohabiting Women", here("Out", "cohabiting_women_final.tex")) 
create_final_table(df, 6, "Single Women", here("Out", "single_women_final.tex"))
```

### Item c
```{r, echo=FALSE}
## Filter for partnered women aged 25-65 (household heads or spouses)
df = sdem %>% 
  filter(e_con %in% c(1,5),        # Marital status: cohabiting (1) or married (5)
         sex == 2,                 # Female
         eda %in% 25:65,           # Age filter 25-65 years
         par_c %in% c(101,201))    # Household head or spouse

## Filter for men (spouses/partners) - no age restriction to capture all male partners
dfM = sdem %>% 
  filter(sex == 1,                 # Male
         par_c %in% c(101,201))    # Household head or spouse

# Clean memory
rm(sdem)
gc()

# Remove households with multiple women (data quality issue)
# This ensures one woman per household for accurate matching
# Aux = df %>% 
#   group_by(cd_a, ent, con, v_sel, n_hog, h_mud, year) %>% 
#   summarise(n = n()) %>% 
#   filter(n > 1) 
# 
# X = paste(df$cd_a, df$ent, df$con, df$v_sel,
#           df$n_hog, df$h_mud, df$year)
# X2 = paste(Aux$cd_a, Aux$ent, Aux$con, Aux$v_sel,
#            Aux$n_hog, Aux$h_mud, Aux$year)
# 
# df = df[!X %in% X2,]
# rm(X, X2, Aux)
df = df %>% 
  group_by(cd_a, ent, con, v_sel, n_hog, h_mud, year) %>% 
  mutate(n = n()) %>% 
  ungroup() %>% 
  filter(n == 1)

# Handle missing values for women's housework variables
# p11_h2/p11_m2: hours/minutes for household chores (cooking, cleaning)
# p11_h5/p11_m5: hours/minutes for care activities (before 2013)
# p11_h7/p11_m7: hours/minutes for care activities (2013 onwards)
df$p11_h2 = ifelse(is.na(df$p11_h2), 0, df$p11_h2)
df$p11_h5 = ifelse(is.na(df$p11_h5), 0, df$p11_h5)
df$p11_h7 = ifelse(is.na(df$p11_h7), 0, df$p11_h7)
df$p11_m2 = ifelse(is.na(df$p11_m2), 0, df$p11_m2)
df$p11_m5 = ifelse(is.na(df$p11_m5), 0, df$p11_m5)
df$p11_m7 = ifelse(is.na(df$p11_m7), 0, df$p11_m7)

# Filter out "don't know" responses (code 98)
df = df %>% 
  filter(p11_h2 != 98, p11_h5 != 98, p11_h7 != 98)

# Convert "no response" codes (99) to zero
df$p11_h2 = ifelse(df$p11_h2 == 99 | is.na(df$p11_h2), 0, df$p11_h2)
df$p11_h5 = ifelse(df$p11_h5 == 99 | is.na(df$p11_h5), 0, df$p11_h5)
df$p11_h7 = ifelse(df$p11_h7 == 99 | is.na(df$p11_h7), 0, df$p11_h7)

# Calculate total weekly household chores time for women (in minutes)
# Survey structure changed in 2013, so we use different variables by year
df = df %>% 
  mutate(hhChores_W = case_when(
    year < 2013 ~ p11_h2*60 + p11_m2 + p11_h5*60 + p11_m5,
    year >= 2013 ~ p11_h2*60 + p11_m2 + p11_h7*60 + p11_m7,
    T ~ 0
  ))
gc()
# Remove households with multiple men (data quality issue)
# Aux = dfM %>% 
#   group_by(cd_a, ent, con, v_sel, n_hog, h_mud, year) %>% 
#   summarise(n = n()) %>% 
#   filter(n > 1) 
# 
# X = paste(dfM$cd_a, dfM$ent, dfM$con, dfM$v_sel,
#           dfM$n_hog, dfM$h_mud, dfM$year)
# X2 = paste(Aux$cd_a, Aux$ent, Aux$con, Aux$v_sel,
#            Aux$n_hog, Aux$h_mud, Aux$year)
# 
# dfM = dfM[!X %in% X2,]
# rm(X, X2, Aux)
dfM = dfM %>% 
  group_by(cd_a, ent, con, v_sel, n_hog, h_mud, year) %>% 
  mutate(n = n()) %>% 
  ungroup() %>% 
  filter(n == 1)

# Handle missing values for men's housework variables
dfM$p11_h2 = ifelse(is.na(dfM$p11_h2), 0, dfM$p11_h2)
dfM$p11_h5 = ifelse(is.na(dfM$p11_h5), 0, dfM$p11_h5)
dfM$p11_h7 = ifelse(is.na(dfM$p11_h7), 0, dfM$p11_h7)
dfM$p11_m2 = ifelse(is.na(dfM$p11_m2), 0, dfM$p11_m2)
dfM$p11_m5 = ifelse(is.na(dfM$p11_m5), 0, dfM$p11_m5)
dfM$p11_m7 = ifelse(is.na(dfM$p11_m7), 0, dfM$p11_m7)

# Filter out "don't know" responses for men
dfM = dfM %>%
  filter(p11_h2 != 98, p11_h5 != 98, p11_h7 != 98)

# Convert "no response" codes to zero for men
dfM$p11_h2 = ifelse(dfM$p11_h2 == 99 | is.na(dfM$p11_h2), 0, dfM$p11_h2)
dfM$p11_h5 = ifelse(dfM$p11_h5 == 99 | is.na(dfM$p11_h5), 0, dfM$p11_h5)
dfM$p11_h7 = ifelse(dfM$p11_h7 == 99 | is.na(dfM$p11_h7), 0, dfM$p11_h7)

# Calculate total weekly household chores time for men (in minutes)
dfM = dfM %>% 
  mutate(hhChores_M = case_when(
    year < 2013 ~ p11_h2*60 + p11_m2 + p11_h5*60 + p11_m5,
    year >= 2013 ~ p11_h2*60 + p11_m2 + p11_h7*60 + p11_m7,
    T ~ 0
  ))

# Select only necessary variables from men's dataset for merging
dfM = dfM %>%
  select(cd_a, ent, con, v_sel, n_hog, h_mud, year, hhChores_M, Wage, IngOcupYear) %>% 
  rename(WageM = Wage,
         IngOcupYearM = IngOcupYear)
gc()
# Merge women and men datasets by household identifiers
df = df %>%
  left_join(dfM, by = c("cd_a", "ent", "con", "v_sel", "n_hog", "h_mud", "year"))

# Remove observations without male household chores data (incomplete couples)
df = df[!is.na(df$hhChores_M),]
gc()

# Create the main plot showing evolution of housework ratio over time
# This directly answers Question 1(c)
housework_ratio_plot = df %>% 
  group_by(year, e_con) %>% 
  summarise(Ratio = mean(hhChores_W/(hhChores_M + hhChores_W), na.rm = T)) %>%
  mutate(marital_status = case_when(
    e_con == 1 ~ "Cohabiting",
    e_con == 5 ~ "Married"
  )) %>%
  ggplot(mapping = aes(year, Ratio, col = as.factor(marital_status))) +
  geom_point() +
  geom_line() + 
  theme_bw() +
  labs(
    # title = "Weekly Housework Hours by Gender and Marital Status",
    # subtitle = "Partnered Women Aged 25-65 and Their Spouses",
    y = "Ratio",
    x = "Year",
    color = "Marital Status"
  ) + 
  scale_x_continuous(breaks = unique(df$year))

# Create supplementary plot showing absolute hours by gender and marital status
absolute_hours_plot = df %>% 
  group_by(year, e_con) %>% 
  summarise(childCare = mean(p11_h2, na.rm = T)) %>%
  mutate(marital_status = case_when(
    e_con == 1 ~ "Cohabiting",
    e_con == 5 ~ "Married"
  )) %>%
  ggplot(mapping = aes(year, childCare, col = as.factor(marital_status))) +
  geom_point() +
  geom_line() + 
  theme_bw() +
  labs(
    # title = "Weekly Housework Hours by Gender and Marital Status",
    # subtitle = "Partnered Women Aged 25-65 and Their Spouses",
    y = "Hours Child Care",
    x = "Year",
    color = "Marital Status"
  ) + 
  scale_x_continuous(breaks = unique(df$year))

# Print summary statistics for interpretation
cat("Summary Statistics for Housework Ratio Analysis:\n")
cat("==============================================\n\n")

# Overall summary by marital status
summary_stats = df %>%
  group_by(e_con) %>%
  summarise(
    marital_status = first(case_when(e_con == 1 ~ "Cohabiting", e_con == 5 ~ "Married")),
    n_couples = n(),
    mean_ratio = round(mean(housework_ratio, na.rm = TRUE), 2),
    median_ratio = round(median(housework_ratio, na.rm = TRUE), 2),
    women_avg_hours = round(mean(hhChores_W / 60, na.rm = TRUE), 1),
    men_avg_hours = round(mean(hhChores_M / 60, na.rm = TRUE), 1),
    .groups = 'drop'
  )

print(summary_stats)

# Clean up intermediate objects
rm(dfM, housework_trends)
gc()

# Saving the graphs
ggsave(here("Out", "plot2.png"), housework_ratio_plot, width = 12, height = 12, dpi=600)
ggsave(here("Out", "plot3.png"), absolute_hours_plot, width = 12, height = 12, dpi=600)
```

### Item d
```{r, echo=FALSE}
## Importing databases
VivEni = read_csv(here("Data/Raw", "viviendas.csv"))
ConcEni = read_csv(here("Data/Raw", "concentradohogar.csv"))
PobEni = read_csv(here("Data/Raw", "poblacion.csv"))
GasEni = read_csv(here("Data/Raw", "gastoshogar.csv"))

# Get Home Ownership status (exactly as original)
VivEni = VivEni %>% 
  filter(tenencia != 6) %>%
  mutate(homeownership = ifelse(tenencia %in% c(1,2,5), 0, 1)) %>%
  select(folioviv, homeownership)

# Extract specific expenditure categories (exactly as original)
GasEni = GasEni %>% 
  mutate(health_ins = ifelse(clave %in% c("J070", "J071", "J072"),
                             gasto_tri, 0),
         other_ins  = ifelse(clave %in% c("N008", "N009"),
                             gasto_tri, 0),
         child_care = ifelse(clave %in% c("E012"),
                             gasto_tri, 0),
         child_exp  = ifelse(!is.na(clave),
                             as.integer(clave == "E012"), NA),
         day_care = ifelse(clave %in% c("E008"),
                           gasto_tri, 0),
         day_exp  = ifelse(!is.na(clave),
                           as.integer(clave == "E008"), NA))

# Aggregate household expenses by household (exactly as original)
GasH = GasEni %>% 
  group_by(folioviv, foliohog) %>%
  summarise(health_ins = sum(health_ins, na.rm = T),
            other_ins  = sum(other_ins, na.rm = T),
            child_care = sum(child_care, na.rm = T),
            child_exp  = sum(child_exp, na.rm = T),
            day_care   = sum(day_care, na.rm = T),
            day_exp    = sum(day_exp, na.rm = T))

# Handle missing values in worked hours (exactly as original)
PobEni = PobEni %>% 
  mutate(hor_1 = ifelse(is.na(hor_1), 0, hor_1),
         min_1 = ifelse(is.na(min_1), 0, min_1))

# Create household-level demographic and work variables (exactly as original)
PobH = PobEni %>% 
  group_by(folioviv, foliohog) %>%
  summarise(age_hhead = edad[which(parentesco == 101)],
            age_hspouse = first(edad[parentesco %in% c(201,202)]),
            married = edo_conyug[which(parentesco == 101)],
            sex_hhead = sexo[which(parentesco == 101)],
            sex_hspouse = sexo[which(parentesco %in% c(201,202))],
            hworked_hhead = 
              hor_1[which(parentesco == 101)]*60 + min_1[which(parentesco == 101)],
            hworked_hspouse = 
              hor_1[which(parentesco %in% c(201,202))]*60 + min_1[which(parentesco %in% c(201,202))],
            sch_hhead = nivelaprob[which(parentesco == 101)],
            sch_hspouse = nivelaprob[which(parentesco %in% c(201,202))])

# Join all datasets (exactly as original)
ConcEni = ConcEni %>% 
  left_join(VivEni, by = c("folioviv"))

ConcEni = ConcEni %>% 
  left_join(GasH, by = c("folioviv", "foliohog"))

ConcEni = ConcEni %>% 
  left_join(PobH, by = c("folioviv", "foliohog"))

# Filter for nuclear households only (exactly as original)
ConcEni = ConcEni %>% 
  filter(clase_hog == 2)

# Create expense and income variables (exactly as original)
ConcEni = ConcEni %>% 
  mutate(
    nonlabor_inc = (ing_cor - ingtrab)/3,
    food =  ali_dentro + ali_fuera,
    transportation = transporte,
    healthService = salud,
    utilities = pred_cons + agua + energia,
    education = educacion,
    houseKeeping = cuidados,
    rent = ifelse(homeownership == 0, alquiler, estim_alqu),
    insurance = health_ins + other_ins,
    health_ins = health_ins,
    home_ins = other_ins,
    childc = day_care + child_care
  ) %>%
  select(folioviv, foliohog, age_hhead, age_hspouse, married,
         homeownership, sex_hhead, sex_hspouse, hworked_hhead,
         hworked_hspouse, sch_hhead, sch_hspouse,
         nonlabor_inc, ingtrab, food, transportation,
         healthService, utilities, education, houseKeeping,
         rent, insurance, health_ins, home_ins, childc)

# Data cleaning steps (exactly as original)
ConcEni = ConcEni %>% 
  filter(!is.na(homeownership))

ConcEni = ConcEni %>% 
  filter(nonlabor_inc < quantile(nonlabor_inc, 0.99, na.rm = TRUE),
         ingtrab < quantile(ingtrab, 0.99, na.rm = TRUE))

ConcEni = ConcEni %>%
  mutate(
    total_consumption = (food + transportation +
                           healthService + houseKeeping + health_ins +
                           home_ins + utilities + childc + education + rent) / 3,
    hworked_hhead = hworked_hhead/60,
    hworked_hspouse = hworked_hspouse/60
  )

# Create separate datasets by marital status (exactly as original)
dfMM = ConcEni %>% filter(married == 2)  # Married couples
dfCM = ConcEni %>% filter(married == 1)  # Cohabiting couples

# Define variables for summary statistics (exactly as original)
vars_to_summarise = c("food", "transportation", "healthService", 
                      "utilities", "education", "houseKeeping", "rent", 
                      "health_ins", "home_ins", "childc")

# Create comprehensive summary table for married couples
tabela_married = dfMM %>%
  select(all_of(vars_to_summarise)) %>%
  summarise(across(everything(), 
                   list(Mean = ~round(mean(.x, na.rm = TRUE), 2),
                        Median = ~round(median(.x, na.rm = TRUE), 2)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  separate(variable, into = c("Category", "Statistic"), sep = "_(?=[^_]*$)") %>%
  pivot_wider(names_from = Statistic, values_from = value) %>%
  mutate(Category = case_when(
    Category == "food" ~ "Food",
    Category == "transportation" ~ "Transportation", 
    Category == "healthService" ~ "Health Services",
    Category == "houseKeeping" ~ "Housekeeping Services",
    Category == "health_ins" ~ "Health Insurance",
    Category == "home_ins" ~ "Home Insurance",
    Category == "utilities" ~ "Utilities",
    Category == "childc" ~ "Child Care",
    Category == "education" ~ "Education",
    Category == "rent" ~ "Rent"
  ))

# Create comprehensive summary table for cohabiting couples  
tabela_cohabiting = dfCM %>%
  select(all_of(vars_to_summarise)) %>%
  summarise(across(everything(), 
                   list(Mean = ~round(mean(.x, na.rm = TRUE), 2),
                        Median = ~round(median(.x, na.rm = TRUE), 2)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  separate(variable, into = c("Category", "Statistic"), sep = "_(?=[^_]*$)") %>%
  pivot_wider(names_from = Statistic, values_from = value) %>%
  mutate(Category = case_when(
    Category == "food" ~ "Food",
    Category == "transportation" ~ "Transportation", 
    Category == "healthService" ~ "Health Services",
    Category == "houseKeeping" ~ "Housekeeping Services",
    Category == "health_ins" ~ "Health Insurance",
    Category == "home_ins" ~ "Home Insurance",
    Category == "utilities" ~ "Utilities",
    Category == "childc" ~ "Child Care",
    Category == "education" ~ "Education",
    Category == "rent" ~ "Rent"
  ))

# Load required packages for LaTeX export
library(xtable)

# Export married couples table to LaTeX
married_latex = xtable(tabela_married,
                       caption = "Mean and Median Expenditures for Households Headed by Married Couples",
                       label = "tab:married_expenditures",
                       digits = 2)

print(married_latex,
      type = "latex",
      file = here("Out", "married_couples_expenditures.tex"),
      include.rownames = FALSE,
      caption.placement = "top",
      table.placement = "htbp")

# Export cohabiting couples table to LaTeX
cohabiting_latex = xtable(tabela_cohabiting,
                          caption = "Mean and Median Expenditures for Households Headed by Cohabiting Couples", 
                          label = "tab:cohabiting_expenditures",
                          digits = 2)

print(cohabiting_latex,
      type = "latex", 
      file = here("Out", "cohabiting_couples_expenditures.tex"),
      include.rownames = FALSE,
      caption.placement = "top",
      table.placement = "htbp")

# Display summary table in console (exactly as original)
print("Married Couples Summary:")
print(tabela_married)
print("Cohabiting Couples Summary:")
print(tabela_cohabiting)

# Print summary information
cat("Summary Statistics Generated:\n")
cat("============================\n")
cat("Number of married couple households:", nrow(dfMM), "\n")
cat("Number of cohabiting couple households:", nrow(dfCM), "\n")
cat("\nLaTeX files generated:\n")
cat("- married_couples_expenditures.tex\n")
cat("- cohabiting_couples_expenditures.tex\n")

# Clean up environment (exactly as original)
rm(VivEni, GasEni, GasH, PobEni, PobH)
gc()
```

# Question 2


# Question 3


# Question 4
## Item a
### Interior solution
Considering positive work in both period 1 and 2, our Lagrangian is

$$\mathcal{L}=\sum_{t=1}^{3}\beta^{t-1}[\alpha \ln C_t+(1-\alpha)\ln L_t]-\lambda \left [\sum_{t=1}^{3} \frac{C_t}{(1+r)^{t-1}}-A_1-\sum_{t=1}^{2}\frac{w_t(L_0-L_t)}{(1+r)^{t-1}}\right]$$

with CPO's:

$$
\begin{aligned}
\,
[C_t] &: \frac{\alpha\beta^{t-1}}{C_t}-\frac{\lambda}{(1+r)^{t-1}}=0 \Longrightarrow C_t^*=\frac{\alpha[\beta(1+r)]^{t-1}}{\lambda} \\
[L_t] &: \frac{(1-\alpha)\beta^{t-1}}{L_t}-\frac{\lambda w_t}{(1+r)^{t-1}}=0 \Longrightarrow L_t^*=\frac{(1-\alpha)[\beta(1+r)]^{t-1}}{\lambda w_t}
\end{aligned}
$$

Then, the Frischian labor supply function is simple

$$
h_t^F(w_t, \lambda, t)=L_0-L_t^*=L_0-\frac{(1-\alpha)[\beta(1+r)]^{t-1}}{\lambda w_t}
$$

Recall that the Marshallian elasticity accounts for variations in labor supply due to permanent shocks in the path of wage. So, to find the Marshallian labor supply function, we need to substitute $C_t^*$ and $L_t^*$ into the budget constraint and isolate $\lambda$ to find a close expression for the Lagrangean multiplier, i. e.,

$$
\begin{aligned}
\sum_{t=1}^{3} \frac{C_t^*}{(1+r)^{t-1}}-A_1-\sum_{t=1}^{2}\frac{w_t(L_0-L_t^*)}{(1+r)^{t-1}} &= 0 \\
\sum_{t=1}^{3} \frac{1}{(1+r)^{t-1}} \cdot \frac{\alpha[\beta(1+r)]^{t-1}}{\lambda}-A_1-\sum_{t=1}^{2}\left[\frac{w_tL_0}{(1+r)^{t-1}}-\frac{w_t}{(1+r)^{t-1}}\cdot \frac{(1-\alpha)[\beta(1+r)]^{t-1}}{\lambda w_t}\right] &= 0 \\
\sum_{t=1}^{3} \frac{\alpha \beta^{t-1}}{\lambda}-A_1-\sum_{t=1}^{2}\left[\frac{w_tL_0}{(1+r)^{t-1}}-\frac{(1-\alpha)\beta^{t-1}}{\lambda}\right] &= 0 \\
\frac{\alpha}{\lambda}\sum_{t=1}^{3}\beta^{t-1}-A_1-L_0\sum_{t=1}^{2}\frac{w_t}{(1+r)^{t-1}}+\frac{(1-\alpha)}{\lambda}\sum_{t=1}^{2}\beta^{t-1} &= 0
\end{aligned}
$$

As we know

$$
\sum_{t=1}^{T}\beta^{t-1}=\frac{1-\beta^T}{1-\beta}
$$
So, we have

$$
\begin{aligned}
A_1+L_0\sum_{t=1}^{2}\frac{w_t}{(1+r)^{t-1}} &= \frac{1}{\lambda(1-\beta)}[\alpha(1-\beta^3)+(1-\alpha)(1-\beta^2)] \\
\lambda \left[A_1+L_0 \left(w_1+\frac{w_2}{(1+r)}\right)\right] &= \frac{1-\alpha \beta^3+(1-\alpha)\beta^2}{(1-\beta)} \\
\lambda \left [\frac{(1+r)(A_1+L_0 w_1)+w_2}{(1+r)} \right] &= \frac{1-\alpha \beta^3+(1-\alpha)\beta^2}{(1-\beta)} \\
\lambda^* &= \frac{(1+r)[1-\alpha \beta^3+(1-\alpha)\beta^2]}{(1-\beta)[(1+r)(A_1+L_0 w_1)+w_2]}
\end{aligned}
$$
Therefore, the Marshallian labor supply function is:

$$
\begin{aligned}
h_t^M(w_t, \lambda^*, t) &= L_0-\frac{(1-\alpha)[\beta(1+r)]^{t-1}}{\lambda^* w_t} \\
h_t^M(w_t, \lambda^*, t) &= L_0-\frac{\gamma(1-\alpha)[\beta(1+r)]^{t-1}}{w_t}
\end{aligned}
$$

where
$$
\gamma = \frac{(1-\beta)[(1+r)(A_1+L_0 w_1)+w_2]}{(1+r)[1-\alpha \beta^3+(1-\alpha)\beta^2]}
$$

### Corner solution
We have two cases to consider here:

1. **Work only in period 1 ($L_2=L_0$)**

In this case, the budget constraint will be

$$
\sum_{t=1}^{3} \frac{C_t}{(1+r)^{t-1}}=A_1+w_1(L_0-L_1)
$$
Notice that in this case we just need to choose $C_t$ and $L_1$, so the Lagrangean will be

$$
\mathcal{L}=\sum_{t=1}^{3}\beta^{t-1}\alpha \ln C_t+(1-\alpha)\ln L_1 + (\beta+\beta^2)(1-\alpha)L_0 + \lambda\left [A_1 + w_1(L_0-L_1)-\sum_{t=1}^3\frac{C_t}{(1+r)^{t-1}} \right]
$$

Then, from FOC we get and repeating the previous process, we get

$$
\begin{aligned}
h_1^F(w_1, \lambda)&=L_0-\frac{1-\alpha}{\lambda w_1} \\
h_1^M(w_1, \lambda^*, t)&=L_0-\frac{\gamma'(1-\alpha)}{w_1}
\end{aligned}
$$

where here

$$
\gamma'=\frac{(1-\beta)[A_1+w_1L_0]}{1-[1-\alpha(1-\beta^2)]\beta}
$$

2. **Work only in period 2 ($L_1=L_0$)**

The budget constraint in this case is

$$
\sum_{t=1}^{3} \frac{C_t}{(1+r)^{t-1}}=A_1+\frac{w_2(L_0-L_2)}{(1+r)}
$$
The Lagrangean then will be

$$
\mathcal{L}=\sum_{t=1}^{3}\beta^{t-1}\alpha \ln C_t+\beta(1-\alpha)\ln L_2 + (1+\beta^2)(1-\alpha)L_0 + \lambda\left [A_1 + \frac{w_2(L_0-L_2)}{(1+r)}-\sum_{t=1}^3\frac{C_t}{(1+r)^{t-1}} \right]
$$

Repeating the process above yields

$$
\begin{aligned}
h_2^F(w_2, \lambda)&=L_0-\frac{\beta(1-\alpha)(1+r)}{\lambda w_2} \\
h_2^M(w_2, \lambda^*, t)&=L_0-\frac{\gamma''(1-\alpha)}{w_2}
\end{aligned}
$$
where

$$
\gamma''=\frac{(1-\beta[(1+r)A_1+w_2L_0])}{(1+r)[\alpha(1-\beta^3)+(1-\alpha)\beta(1-\beta)]}
$$